Exploration of clustering results for `r params$sample_id`

### 2024-08-14 analyses/cell-type-wilms-tumor-06/notebook-template/01-clustering.Rmd
Update analyses/cell-type-wilms-tumor-06/notebook-template/01-clustering.Rmd
Co-authored-by: Stephanie Spielman <stephanie.spielman@gmail.com>
Add to PR #704

---
title: "Exploration of clustering results"
author: maud-p
date: '`r format(Sys.time(), "%B %d, %Y %H:%M:%S %Z")`'
output:
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: kable
params:
  sample_name: ~
  sample_path: ~
  out_rds_path: ~
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warnings = FALSE,
	error = TRUE
)
```

@@

This a template for an analysis notebook using RMarkdown.

In this notebook, we will set up parameters in the seurat workflow [normalization --> reduction --> clustering] for one Wilms tumor sample (SCPCS000169) of the Wilms Tumor dataset (SCPCP000006).

@@

In order to assess the clustering quality, we look into some marker genes, pathways enrichment and label transfer.

## Sample metadata

```{r paths2}
## to be modified when rendered
filedir <- filelist[i] 
sample <- gsub(paste0(data_dir, "/"), "", filedir)

metadata <- read.table(file.path(data_dir, "/single_cell_metadata.tsv"), sep = "\t", header=TRUE)

metadata[metadata$scpca_sample_id == sample,]
```

@@

[4] We look at specific pathways that we reported in the table marker.sets/Pathways_metadata.csv to check the relevance of the clustering.

[5] We run DElegate::FindAllMarkers2 to find markers of the different clusters and manually check if they do make sense. 
DElegate::FindAllMarkers2 is an improved version of Seurat::FindAllMarkers based on pseudobulk differential expression method.

[6] We perform enrichment analysis of marker genes for each seurat clusters. 
We defined all the genes from the seurat object as the universe and used the MSigDB gene sets.

[7] We plot pca/umap reduction grouping with available annotations (singler_, cellassign_). 
We expect at least immune cells to be correctly label and fall into a few set of clusters.

[ ] Next step, aim for a next PR: We will label transfer (Azimuth) to transfer annotation from the fetal kidney atlas human reference. 
We plot pca/umap reduction grouping with latest labels. 
We expect it to be the most representative of the cell types in the sample.

@@

@@

```{r load}
file <- list.files(filedir)
file <- file[grepl("_processed.rds", file)]

@@

```{r seurat_wf, fig.width=8, fig.height=4, out.width='100%'}
# convert to seurat
srat <- CreateSeuratObject(counts = counts(sce),
                                    assay = "RNA",
                                    project = sample
                           )

@@

### Cell cycle information
```{r cell_cycle, fig.height=20, fig.width=30}
s.genes <- srat@assays$RNA@meta.data$gene_ids[srat@assays$RNA@meta.data$gene_symbol %in% cc.genes$s.genes]
g2m.genes <- srat@assays$RNA@meta.data$gene_ids[srat@assays$RNA@meta.data$gene_symbol %in% cc.genes$g2m.genes]

@@

d2 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "Phase", label = TRUE) + ggtitle("Phase - umap")
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE) + ggtitle("Seurat Cluster - umap")

b <- SCpubr::do_BarPlot(sample = srat,
                         group.by = "Phase",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 3) +
                         ggtitle("% cells")+
                         xlab(file)
d1 + d2 + b + plot_layout(ncol = 3, widths = c(2,2,4))

```{r fig.width=10, fig.height=2.5, out.width='100%'}
 SCpubr::do_ViolinPlot(srat, features = c("S.Score", "G2M.Score"), ncol = 2, group.by = "seurat_clusters", legend.position = "none") 
```

@@

```{r marker_genes, fig.width=12, fig.height=6, out.width='100%'}

@@

#### Malignant markers

```{r fig.width=25, fig.height=5, out.width='100%'}
 SCpubr::do_ViolinPlot(srat, features = rownames(srat)[rownames(srat) %in% (CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class %in% c("malignant")])] , ncol = 9, group.by = "seurat_clusters", legend.position = "none") 
```

##### NCAM1

```{r fig.width=20, fig.height=5, out.width='100%'}
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000149294", pt.size = 0.2) + ggtitle("NCAM1")
v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000149294", ncol = 1, group.by = "seurat_clusters", legend.position = "none")

d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
```

##### COL6A3

```{r fig.width=20, fig.height=5, out.width='100%'}
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000163359", pt.size = 0.2) + ggtitle("COL6A3")
v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000163359", ncol = 1, group.by = "seurat_clusters", legend.position = "none")

d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))

##### IGF2

```{r fig.width=20, fig.height=5, out.width='100%'}
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000167244", pt.size = 0.2) + ggtitle("IGF2")
v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000167244", ncol = 1, group.by = "seurat_clusters", legend.position = "none")

d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
```

##### WT1

```{r fig.width=20, fig.height=5, out.width='100%'}
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000184937", pt.size = 0.2) + ggtitle("WT1")
v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000184937", ncol = 1, group.by = "seurat_clusters", legend.position = "none")

d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
```

##### EPCAM

```{r fig.width=20, fig.height=5, out.width='100%'}
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000119888", pt.size = 0.2) + ggtitle("EPCAM")
v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000119888", ncol = 1, group.by = "seurat_clusters", legend.position = "none")

d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
```

#### Immune markers

```{r fig.width=20, fig.height=5, out.width='100%'}
 SCpubr::do_ViolinPlot(srat, features = rownames(srat)[rownames(srat) %in% (CellType_metadata$ENSEMBL_ID[CellType_metadata$cell_class %in% c("immune")])]
                       , ncol = 6, group.by = "seurat_clusters", legend.position = "none") 
```

##### PTPRC = CD45

```{r fig.width=20, fig.height=5, out.width='100%'}
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000081237", pt.size = 0.2) + ggtitle("PTPRC = CD45")
v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000081237", ncol = 1, group.by = "seurat_clusters", legend.position = "none", pt.size = 2)

d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
```

#### Other markers

```{r fig.width=20, fig.height=5, out.width='100%'}
 SCpubr::do_ViolinPlot(srat, features = rownames(srat)[rownames(srat) %in% (CellType_metadata$ENSEMBL_ID[! CellType_metadata$cell_class %in% c("immune", "malignant")])]
                       , ncol = 6, group.by = "seurat_clusters", legend.position = "none") 
```

##### VWF, endothelial cells

```{r fig.width=20, fig.height=5, out.width='100%'}
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
d2 <- SCpubr::do_FeaturePlot(srat, features = "ENSG00000110799", pt.size = 0.2) + ggtitle("VWF")
v <- SCpubr::do_ViolinPlot(srat, features = "ENSG00000110799", ncol = 1, group.by = "seurat_clusters", legend.position = "none", pt.size = 2)

d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
```

@@

Here, we open the table of marker genes marker-sets/Pathway_metadata.csv.

```{r marker_pathways, fig.width=12, fig.height=6, out.width='100%'}

DT::datatable(Pathway_metadata, caption = ("Pathway_metadata"), 
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',

@@

```{r fig.width=10, fig.height=10, out.width='100%'}

## define genesets
hallmarks <- msigdbr(species = "human", category = "H")

TP53_list = hallmarks %>% 
  filter(gs_name == "HALLMARK_P53_PATHWAY") %>%
  dplyr::distinct(gs_name, gene_symbol, human_ensembl_gene) %>% as.data.frame()

TP53_list <- list(TP53_list$human_ensembl_gene)

srat <- AddModuleScore(srat, features = TP53_list, name = "TP53_score")
```

```{r fig.width=20, fig.height=5, out.width='100%'}
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
d2 <- SCpubr::do_FeaturePlot(srat, features = "TP53_score1", pt.size = 0.2) + ggtitle("TP53 score")
v <- SCpubr::do_ViolinPlot(srat, features = "TP53_score1", ncol = 1, group.by = "seurat_clusters", legend.position = "none")

d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
```

@@

```{r fig.width=10, fig.height=10, out.width='100%'}

DNA_repair = hallmarks %>% 
  filter(gs_name == "HALLMARK_DNA_REPAIR") %>%
  dplyr::distinct(gs_name, gene_symbol, human_ensembl_gene) %>% as.data.frame()

DNA_repair_list <- list(DNA_repair$human_ensembl_gene)

srat <- AddModuleScore(srat, features = DNA_repair_list, name = "DNA_repair_score")
```

```{r fig.width=20, fig.height=5, out.width='100%'}
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
d2 <- SCpubr::do_FeaturePlot(srat, features = "DNA_repair_score1", pt.size = 0.2) + ggtitle("DNA_repair score")
v <- SCpubr::do_ViolinPlot(srat, features = "DNA_repair_score1", ncol = 1, group.by = "seurat_clusters", legend.position = "none")

d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
```

@@

```{r fig.width=10, fig.height=10, out.width='100%'}

## define genesets
c3 <- msigdbr(species = "human", category = "C3", subcategory = "TFT:GTRD")

DROSHA_list = c3 %>% 
  filter(gs_name == "DROSHA_TARGET_GENES") %>%
  dplyr::distinct(gs_name, gene_symbol, human_ensembl_gene) %>% as.data.frame()

DROSHA_list <- list(DROSHA_list$human_ensembl_gene)

srat <- AddModuleScore(srat, features = DROSHA_list, name = "DROSHA_score")
```

```{r fig.width=20, fig.height=5, out.width='100%'}
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
d2 <- SCpubr::do_FeaturePlot(srat, features = "DROSHA_score1", pt.size = 0.2) + ggtitle("DROSHA score")
v <- SCpubr::do_ViolinPlot(srat, features = "DROSHA_score1", ncol = 1, group.by = "seurat_clusters", legend.position = "none")

d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
```

@@

```{r fig.width=10, fig.height=10, out.width='100%'}
DICER1_list = c3 %>% 
  filter(gs_name == "DICER1_TARGET_GENES") %>%
  dplyr::distinct(gs_name, gene_symbol, human_ensembl_gene) %>% as.data.frame()

DICER1_list <- list(DICER1_list$human_ensembl_gene)

srat <- AddModuleScore(srat, features = DICER1_list, name = "DICER1_score")
```

```{r fig.width=20, fig.height=5, out.width='100%'}
d1 <- SCpubr::do_DimPlot(srat, reduction="umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + ggtitle("Seurat Cluster - umap")
d2 <- SCpubr::do_FeaturePlot(srat, features = "DICER1_score1", pt.size = 0.2) + ggtitle("DICER1 score")
v <- SCpubr::do_ViolinPlot(srat, features = "DICER1_score1", ncol = 1, group.by = "seurat_clusters", legend.position = "none")

d1 + d2 + v + plot_layout(ncol = 3, widths = c(1,1,3))
```

@@

@@

#filter the most relevant markers
s.markers <- de_results[de_results$padj < cfg$padj_thershold & de_results$log_fc > cfg$lfc_threshold & de_results$rate1 > cfg$rate1_threshold,]

@@

```{r fig.width=10, fig.height=10, out.width='100%'}
# define background genes = universe for enrichment
background <-  row_metadata$gene_ids

## define genesets
c8 <- msigdbr(species = "human", category = "C8")

msigdbr_hallmarks = hallmarks %>% dplyr::distinct(gs_name, ensembl_gene) %>% as.data.frame()
msigdbr_c3 = c3 %>% dplyr::distinct(gs_name, ensembl_gene) %>% as.data.frame()
msigdbr_c8 = c8 %>% dplyr::distinct(gs_name, ensembl_gene) %>% as.data.frame()
```

@@

```{r fig.width=25, fig.height=10, out.width='100%'}
tmp_H <- NULL
for(i in unique(srat$seurat_clusters)){
signature = s.markers$feature[s.markers$group1 == i]
ego_module <- enricher(gene = signature, universe = background, TERM2GENE = msigdbr_hallmarks)
tmp <- NULL

  if(!is.null(ego_module)){
    if(dim(ego_module)[1]>0){
    b_H <- barplot(ego_module, showCategory = 15, font.size = 20, label_format = 40)  
    tmp <- b_H$data
    tmp$set <- "Hallmarks"
    tmp$cluster <- i
    }
  }
tmp_H <- rbind(tmp_H, tmp)
}

ggplot(tmp_H, aes(Count, ID)) +
    geom_bar(stat = "identity", aes(fill=cluster))+

  theme_classic()+ 
    geom_text(
        aes(label = (paste( "Padj = ", round(p.adjust,2)))),
        color = "black",
        size = 3,
        hjust=1,
        position = position_dodge(0.5)
    )+
   theme(text = element_text(size=14))+
  facet_wrap(facets=c("cluster"), ncol = length(unique(tmp_H$cluster)))
```

@@

```{r fig.width=25, fig.height=10, out.width='100%'}

tmp_H <- NULL
for(i in unique(srat$seurat_clusters)){
signature = s.markers$feature[s.markers$group1 == i]
ego_module <- enricher(gene = signature, universe = background, TERM2GENE = msigdbr_c3)
tmp <- NULL

  if(!is.null(ego_module)){
    if(dim(ego_module)[1]>0){
    b_H <- barplot(ego_module, showCategory = 15, font.size = 20, label_format = 40)  
    tmp <- b_H$data
    tmp$ID <- factor(tmp$ID, levels = c(tmp$ID[order(tmp$Count)]))
    tmp$set <- "c8"
    tmp$cluster <- i
    }
  }
tmp_H <- rbind(tmp_H, tmp[1:5,])
}

ggplot(tmp_H, aes(Count, ID)) +
    geom_bar(stat = "identity", aes(fill=cluster))+

  theme_classic()+ 
    geom_text(
        aes(label = (paste( "Padj = ", round(p.adjust,2)))),
        color = "black",
        size = 3,
        hjust=1,
        position = position_dodge(0.5)
    )+
   theme(text = element_text(size=14))+
  facet_wrap(facets=c("cluster"), ncol = length(unique(tmp_H$cluster)))
```

Here to check of we catch and MIR pattern (DROSHA; DICER1; other?)

@@

```{r fig.width=30, fig.height=15, out.width='100%'}

tmp_H <- NULL
for(i in unique(srat$seurat_clusters)){
signature = s.markers$feature[s.markers$group1 == i]
ego_module <- enricher(gene = signature, universe = background, TERM2GENE = msigdbr_c8)
tmp <- NULL

  if(!is.null(ego_module)){
    if(dim(ego_module)[1]>0){
    b_H <- barplot(ego_module, showCategory = 15, font.size = 20, label_format = 40)  
    tmp <- b_H$data
    tmp$ID <- factor(tmp$ID, levels = c(tmp$ID[order(tmp$Count)]))
    tmp$set <- "c8"
    tmp$cluster <- i
    }
  }
tmp_H <- rbind(tmp_H, tmp[1:5,])
}

ggplot(tmp_H, aes(Count, ID)) +
    geom_bar(stat = "identity", aes(fill=cluster))+

  theme_classic()+ 
    geom_text(
        aes(label = (paste( "Padj = ", round(p.adjust,2)))),
        color = "black",
        size = 3,
        hjust=1,
        position = position_dodge(0.5)
    )+
   theme(text = element_text(size=14))+
  facet_wrap(facets=c("cluster"), ncol = length(unique(tmp_H$cluster)))
```

@@

```{r fig.width=15, fig.height=10, out.width='100%'}

d2 <-DimPlot(srat, group.by = "singler_celltype_annotation", reduction = "umap", label = TRUE, repel = TRUE) + NoLegend()

DT::datatable(table(srat$seurat_clusters, srat$singler_celltype_annotation), caption = ("table of SingleR annotation per seurat clusters"), 
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',

d3 <- SCpubr::do_BarPlot(sample = srat,
                         group.by = "singler_celltype_annotation",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 3) +
                         ggtitle("% cells")+
                         xlab(file)

d2|d3
```

@@

```{r fig.width=15, fig.height=10, out.width='100%'}

d2 <-DimPlot(srat, group.by = "cellassign_celltype_annotation", reduction = "umap", label = TRUE, repel = TRUE) + NoLegend()

DT::datatable(table(srat$seurat_clusters, srat$cellassign_celltype_annotation)
, caption = ("table of cellassign annotation per seurat clusters"), 
              extensions = 'Buttons', 
              options = list(  dom = 'Bfrtip',

d3 <- SCpubr::do_BarPlot(sample = srat,
                         group.by = "cellassign_celltype_annotation",
                         split.by = "seurat_clusters",
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 3) +
                         ggtitle("% cells")+
                         xlab(file)

d2|d3
```

@@

To be added in a next PR.

@@

- endotheliúm
- stroma
- fetal nephron 
- immune

@@

## Next step

-[x] We will adapt this template to be render on all the 40 Wilms tumor samples of the dataset.

-[ ] We will save for each sample the rds file

-[ ] We will run inferCNV for each sample to decide the malignant/normal status of some stroma and epithelial cluster, and confirm the blastema annotation.

The next step will provide us a better understanding of the entire cohort. We will then have to set up a strategy to annotate each sample. Open questions are:

[ ] should we annotate single cell                      
or   
[] consider applying similar annotations to all cells in a cluster?

[ ] manual annotation of each cluster / each patient    
or   
[] automated annotation using some threshold?

>>>>

---
title: "Exploration of clustering results for `r params$sample_id`"
author: maud-p
date: '`r format(Sys.time(), "%B %d, %Y %H:%M:%S %Z")`'
output:
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: kable
    self_contained: yes
    mode: selfcontained 
params:
  current: "2024-07-08"
  scpca_project_id: "SCPCP000006"
  sample_id: "SCPCS000176"
  seed: 12345
  padj_thershold: 0.05  # Threshold for adjusted p-value
  lfc_threshold: 1      # Threshold for log fold change
  rate1_threshold: 0.5  # Threshold for rate (percentage) of cells expressing the marker genes in group1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warnings = FALSE
)
```

@@

This a template for an analysis notebook using `RMarkdown`.

In this notebook, we will perform the `Seurat` workflow [normalization --> reduction --> clustering] and explore the clustering results.

@@

In order to assess the clustering quality, we look into some marker genes, pathways enrichment and label transfer.
This approach would provide us a rapid idea of the quality of the clustering:

- in the case each cluster do have a set of specific marker genes, we could expect each of the cluster to have at least a different phenotype,
- if marker genes are mostly shared between clusters, we might have overclustered,
- if no marker genes are found, the quality of cells in the clusters might be impaired (high mitochondrial content? ribosomal content?)

@@

[4] We look at specific pathways to check the relevance of the clustering.

[5] We run `DElegate::FindAllMarkers2` to find markers of the different clusters and manually check if they do make sense. 
`DElegate::FindAllMarkers2` is an improved version of `Seurat::FindAllMarkers` based on pseudobulk differential expression method.

[6] We perform enrichment analysis of marker genes for each Seurat clusters. 
We defined all the genes from the Seurat object as the universe and used the MSigDB gene sets.

[7] We plot `pca`/`umap` reduction grouping with available annotations (`singler_`, `cellassign_`). 
We expect at least immune cells to be correctly label and fall into a few set of clusters.

[ ] Next step, aim for a next PR: We will transfer annotations from two human fetal (kidney) references (Stewart et al., Cao et al.). 
We plot pca/umap reduction grouping with latest labels. 
We expect it to be the most representative of the cell types in the sample.

### Packages

Load required packages in the following chunk, if needed.
Do not install packages here; only load them with the `library()` function.

```{r packages, message=FALSE, warning=FALSE}
library(SingleCellExperiment)

library("Seurat")

#library(Azimuth)            # Will be required once we add the labl transfer from Azimuth fetal reference
#library(SeuratData)         # Will be required once we add the labl transfer from Azimuth fetal reference

library(viridis)            # for plotting with viridis colors
library(SCpubr)             # for plotting 
library(tidyverse)
library(patchwork)

library(msigdbr)
library(enrichplot)
library(clusterProfiler)

library(org.Hs.eg.db)

set.seed(params$seed)
```

### Base directories

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", params$current, params$scpca_project_id)

# The path to this module
module_base <- file.path(repository_base, "analyses", "cell-type-wilms-tumor-06")
```

@@

### Functions

Here we defined function that will be used multiple time all along the notebook.

#### Visualize seurat clusters and markers genes

For a Seurat object `object`and a features `features`, the function `visualize_feature` will plot `FeaturePlot` and `ViolinPlot`

- `object` is the Seurat object

- `features` the gene or quantitative value to be plotted

- `group.by` is the metadata used for grouping the violin plots

```{r }

visualize_feature <- function(object, features, group.by = "seurat_clusters"){

                  feature_symbol = AnnotationDbi::select(org.Hs.eg.db, 
                                keys=features, 
                                columns="SYMBOL", 
                                keytype="ENSEMBL")

                  d <- SCpubr::do_FeaturePlot(object, 
                                              features = feature_symbol$ENSEMBL, 
                                              pt.size = 0.2, 
                                              legend.width = 0.5, 
                                              legend.length = 5, 
                                              legend.position = "right") + ggtitle(feature_symbol$SYMBOL)

                  b <- SCpubr::do_ViolinPlot(srat, 
                                             features = feature_symbol$ENSEMBL, 
                                             ncol = 1, 
                                             group.by = group.by, 
                                             legend.position = "none") + ylab(feature_symbol$SYMBOL)

                  return(d + b + plot_layout(ncol = 2, widths = c(2,4)))

}

#### Visualize seurat clusters and metadata

For a Seurat object `object`and a metadata `metadata`, the function `visualize_metadata` will plot `FeaturePlot` and `BarPlot`

- `object` is the Seurat object

- `metadata` the gene or quantitative value to be plotted

- `group.by` is the metadata used for grouping the violin plots

```{r }

visualize_metadata <- function(object, meta, group.by){

  if(class(object@meta.data[,meta]) == "numeric"){
     d <- SCpubr::do_FeaturePlot(object, 
                                              features = meta, 
                                              pt.size = 0.2, 
                                              legend.width = 0.5, 
                                              legend.length = 5, 
                                              legend.position = "right") + ggtitle(meta)

    b <- SCpubr::do_ViolinPlot(srat, 
                                             features = meta, 
                                             ncol = 1, 
                                             group.by = group.by, 
                                             legend.position = "none")

   return(d + b + plot_layout(ncol = 2, widths = c(2,4))) 
  }

  else{

  d <- SCpubr::do_DimPlot(object, reduction="umap", group.by = group.by, label = TRUE, repel = TRUE) + ggtitle(paste0(meta," - umap")) + theme(text=element_text(size=18))

  b <- SCpubr::do_BarPlot(sample = object,
                         group.by = meta,
                         split.by = group.by,
                         position = "fill",
                         font.size = 10,
                         legend.ncol = 3) +
                         ggtitle("% cells")+
                         xlab(file) +
                         theme(text=element_text(size=18))
  return(d + b + plot_layout(ncol = 2, widths = c(2,4)))
  }

}

#### Calculate `ModuleScore` from a `MSigDB` dataset

For a Seurat object `object`, the function `MSigDB_score` will calculate a score using `AddModuleScore()` function for a MSigDB gene set `gs_name` in the category `category`

- `object` is the Seurat object to calculate the score. 
Score will be added in the metadata of this object.

- `category` is a MSigDB collection (https://www.gsea-msigdb.org/gsea/msigdb/collections.jsp).
values in `c("H", "C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8")`

- `gs_name` is the name of a MSigDB gene set, e.g. `"HALLMARK_P53_PATHWAY"`

- `name` is the name of the module

```{r }

MSigDB_score <- function(object, category, gs_name, name){

        set <- msigdbr(species = "human", category = category)

        set_list <- set %>% 
                    dplyr::filter(gs_name == gs_name) %>%
                    dplyr::distinct(gs_name, gene_symbol, human_ensembl_gene) %>% as.data.frame()

        set_list <- list(set_list$human_ensembl_gene)

        object <- AddModuleScore(object, features = set_list, name = name)

return(object)
}

#### Enrichment analysis

`Enrichment_plot`aim to perform enrichment of the marker genes for each of the Seurat clusters and summarize the results in a `dotplot`.

- `category` is the MSigDB category or collection to be used
values in `c("H", "C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8")`

- `signatures` is a list of marker genes per cluster

- `background`is the universe used for enrichment

```{r }

Enrichment_plot <- function(category, signatures, background){

    ## define genesets
    gene_set <- msigdbr(species = "human", category = category)
    msigdbr_set <-  gene_set %>% dplyr::distinct(gs_name, ensembl_gene) %>% as.data.frame()

    cclust<-compareCluster(geneCluster = signatures, 
               fun = enricher,
               TERM2GENE = msigdbr_set,
               universe=background)
    d <- dotplot(cclust,showCategory=15) + scale_y_discrete(labels=function(x) str_wrap(x, width=40))
    return(d)  
}

@@

```{r load}
filelist <- list.files(data_dir, full.names = TRUE)

# select the 40 Wilms tumor single nucleus folders only
filedir <- filelist[grepl(params$sample_id, filelist)]

file <- list.files(filedir)
file <- file[grepl("_processed.rds", file)]

@@

```{r seurat_wf}
# convert to seurat
srat <- CreateSeuratObject(counts = counts(sce),
                                    assay = "RNA",
                                    project = params$sample_id
                           )

@@

### Cell cycle information
```{r cell_cycle}
s.genes <- srat@assays$RNA@meta.data$gene_ids[srat@assays$RNA@meta.data$gene_symbol %in% cc.genes$s.genes]
g2m.genes <- srat@assays$RNA@meta.data$gene_ids[srat@assays$RNA@meta.data$gene_symbol %in% cc.genes$g2m.genes]

@@

```{r fig.width=10, fig.height=4, out.width='100%'}
visualize_metadata(srat, meta = "Phase", group.by = "seurat_clusters")
```

```{r fig.width=8, fig.height=2.5, out.width='100%'}
visualize_metadata(srat, meta  = "S.Score", group.by = "seurat_clusters")
visualize_metadata(srat, meta = "G2M.Score", group.by = "seurat_clusters")
```

@@

```{r marker_genes, fig.width=12, fig.height=6, out.width='100%'}
CellType_metadata <- read_csv(file.path(module_base, "marker-sets", "CellType_metadata.csv"))

@@

```{r fig.width=8, fig.height=2.5, out.width='100%'}
for(feature in CellType_metadata$ENSEMBL_ID){
  if(feature %in% rownames(srat)){
    tmp <-  visualize_feature(srat, features = feature, group.by = "seurat_clusters")
    print(tmp)
  }
}
```

@@

@@

```{r fig.height=2.5, fig.width=8, warning=FALSE, out.width='100%'}
srat <- MSigDB_score(object = srat , category = "H", gs_name = "HALLMARK_P53_PATHWAY", name = "TP53_score")
visualize_metadata(srat, meta = "TP53_score1", group.by = "seurat_clusters" )
```

@@

```{r fig.height=2.5, fig.width=8, warning=FALSE, out.width='100%'}
srat <- MSigDB_score(object = srat , category = "H", gs_name = "HALLMARK_DNA_REPAIR", name = "DNA_repair_score")
visualize_metadata(srat, meta = "DNA_repair_score1", group.by = "seurat_clusters" )
```

@@

```{r fig.height=2.5, fig.width=8, warning=FALSE, out.width='100%'}
srat <- MSigDB_score(object = srat , category = "C3", gs_name = "DROSHA_TARGET_GENES", name = "DROSHA_score")
visualize_metadata(srat, meta = "DROSHA_score1", group.by = "seurat_clusters" )
```

@@

```{r fig.height=2.5, fig.width=8, warning=FALSE, out.width='100%'}
srat <- MSigDB_score(object = srat , category = "C3", gs_name = "DICER1_TARGET_GENES", name = "DICER1_score")
visualize_metadata(srat, meta = "DICER1_score1", group.by = "seurat_clusters" )
```

@@

Of note, we won't use it for annotation, this is just here to get an idea!

@@

#filter the most relevant markers
s.markers <- de_results[de_results$padj < params$padj_thershold & de_results$log_fc > params$lfc_threshold & de_results$rate1 > params$rate1_threshold,]

@@

```{r enrichment}
# define background genes = universe for enrichment
background <-  row_metadata$gene_ids

# Define gene signature per cluster
signatures <- list()
for( i in unique(s.markers$group1)){
  signatures[[paste0("cluster ",i)]] <- s.markers$feature[s.markers$group1 == i]
}
```

@@

```{r fig.width=15, fig.height=8, out.width='100%'}
tryCatch(
  expr =  {
    Enrichment_plot(category = "H", signatures = signatures, background = background)
  },
  error = function(e) {
    print("No enrichment")
  }
)
```

@@

```{r fig.width=20, fig.height=25, out.width='100%'}
tryCatch(
  expr =  {
Enrichment_plot(category = "C3", signatures = signatures, background = background)
  },
  error = function(e) {
    print("No enrichment")
  }
)
```
Here to check of we catch and MIR pattern (DROSHA; DICER1; other?)

@@

```{r fig.width=20, fig.height=25, out.width='100%'}
tryCatch(
  expr =  {
Enrichment_plot(category = "C8", signatures = signatures, background = background)
  },
  error = function(e) {
    print("No enrichment")
  }
)
```

@@

```{r fig.width=15, fig.height=10, out.width='100%'}
visualize_metadata(srat, meta = "singler_celltype_annotation", group.by = "seurat_clusters")
```

@@

```{r fig.width=15, fig.height=10, out.width='100%'}
visualize_metadata(srat, meta = "cellassign_celltype_annotation", group.by = "seurat_clusters")
```

@@

To be added in a next PR. Look at opened PR #706

@@

- endothelium
- stroma
- fetal nephron 
- immune

@@
